---
title: "Assignment 1 - Language Development in ASD - part 3"
author: "Riccardo Fusaroli"
date: "August 10, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Welcome to the third exciting part of the Language Development in ASD exercise

In this exercise we will delve more in depth with different practices of model comparison and model selection, by first evaluating your models from last time, then learning how to cross-validate models and finally how to systematically compare models.

N.B. There are several datasets for this exercise, so pay attention to which one you are using!

1. The (training) dataset from last time (the awesome one you produced :-) ).
2. The (test) datasets on which you can test the models from last time:
* Demographic and clinical data: https://www.dropbox.com/s/ra99bdvm6fzay3g/demo_test.csv?dl=0
* Utterance Length data: https://www.dropbox.com/s/uxtqqzl18nwxowq/LU_test.csv?dl=0
* Word data: https://www.dropbox.com/s/1ces4hv8kh0stov/token_test.csv?dl=0

```{r}
# Load libraries
library(lmtest)
library(lme4)
library(lmerTest)
library(ggplot2)
library(dplyr)
library(MuMIn)
library(car)
library(plyr)
library(stringr)
library(tidyverse)
library(Metrics)
library(modelr)
library(caret)
library(cvTools)
```

### Exercise 1) Testing model performance

How did your models from last time perform? In this exercise you have to compare the results on the training data and on the test data. Report both of them. Compare them. Discuss why they are different.

- recreate the models you chose last time (just write the code again and apply it to Assignment2TrainData1.csv)
```{r}
# Set WD and read data
setwd("C:/Users/Karolina/Desktop/AU/Experimental Methods III/Assignments/ExpM3-A2")
d_train = read.csv("A1_train.csv")

# Model
m = lmer(CHI_MLU ~ 1 + VISIT + Diagnosis + verbalIQ + types_CHI + (1+VISIT|SUBJ), data=d_train)
summary(m)

# R2
r.squaredGLMM(m)
```
- calculate performance of the model on the training data: root mean square error is a good measure. (Tip: google the functions rmse() and predict() )
```{r}
# rmse closer to 0 means good results

# Prediction
pred = predict(m,d_train)
Metrics::rmse(pred,d_train$CHI_MLU)
```
- create the test dataset (apply the code from assignment 1 part 1 to clean up the 3 test datasets)
```{r}
# Read datasets
demo = read.csv("demo_test.csv")
LU = read.csv("LU_test.csv")
token = read.csv("token_test.csv")

# Rename Child.ID
demo = plyr::rename(demo, c(Child.ID = "SUBJ", Visit = "VISIT"))

# Homogeneize VISIT
LU$VISIT = str_extract(LU$VISIT,"\\d")  
token$VISIT = str_extract(token$VISIT,"\\d") 

# Homogeneize SUBJ
demo$SUBJ = gsub("\\.", "", demo$SUBJ)
LU$SUBJ = gsub("\\.", "", LU$SUBJ)
token$SUBJ = gsub("\\.", "", token$SUBJ)

# Rename verbal and nonverbal IQ
demo = plyr::rename(demo, c(MullenRaw="nonVerbalIQ"))
demo = plyr::rename(demo, c(ExpressiveLangRaw="verbalIQ"))

# Create subsets
demo_test = select(demo, SUBJ, VISIT, Ethnicity, Diagnosis, Gender, Age, ADOS,  nonVerbalIQ, verbalIQ)
LU_test = select(LU, SUBJ, VISIT, MOT_MLU, MOT_LUstd, CHI_MLU, CHI_LUstd)
token_test = select(token, SUBJ, VISIT, types_MOT, types_CHI, tokens_MOT, tokens_CHI)

# Merge
temporal = merge(LU_test, token_test, by=c("SUBJ", "VISIT"))
total = merge(temporal, demo_test, by=c("SUBJ", "VISIT"))

# Select only 1st visits
total_sub= subset(total, VISIT == 1)

# Merge with the old dataset
d_test = merge(total, select(total_sub, SUBJ, ADOS, nonVerbalIQ, verbalIQ), by="SUBJ")

# Delete old columns
d_test = subset(d_test, select = -c(ADOS.x, nonVerbalIQ.x, verbalIQ.x))

# Rename the columns
d_test = plyr::rename(d_test, c(ADOS.y="ADOS"))
d_test = plyr::rename(d_test, c(nonVerbalIQ.y="nonVerbalIQ"))
d_test = plyr::rename(d_test, c(verbalIQ.y="verbalIQ"))

# Turn SUBJ to numeric
d_test$SUBJ = as.numeric(as.factor(d_test$SUBJ))  # as.factor changes names to numbers

# Turn VISIT to numeric
d_test$VISIT = as.numeric(d_test$VISIT)

# Change gender: 1,2 -> M, F
d_test$Gender = factor(d_test$Gender, levels = c("1", "2"), labels = c("M", "F"))

# Change diagnosis: A,B -> ASD, TD
d_test$Diagnosis = factor(d_test$Diagnosis, levels = c("A", "B"), labels = c("ASD", "TD"))

# Save data
write.csv(d_test, file = "A3_test.csv")

```
- test the performance of the models on the test data (Tips: time to reuse "predict()")
```{r}
pred = predict(m,d_test)
Metrics::rmse(pred,d_test$CHI_MLU)
```
Performance of the selected model was calculated on the training dataset (produced in the Assignment 1) and on the test dataset using root mean square error (rmse). Rmse was 0.3056 on the training dataset and 0.5795 on the test dataset.  

- optional: predictions are never certain, can you identify the uncertainty of the predictions? (e.g. google predictinterval())



### Exercise 2) Model Selection via Cross-validation (N.B: ChildMLU!)

One way to reduce bad surprises when testing a model on new data is to train the model via cross-validation. 

In this exercise you have to use cross-validation to calculate the predictive error of your models and use this predictive error to select the best possible model.

- Create the basic model of ChildMLU as a function of Time and Diagnosis (don't forget the random effects!).
```{r}
# Basic model
mb = lmer(CHI_MLU ~ 1 + VISIT + Diagnosis + (1+VISIT|SUBJ), data=d_train)
summary(mb)
```
- Make a cross-validated version of the model. (Tips: google the function "createFolds";  loop through each fold, train a model on the other folds and test it on the fold)
```{r}
# Create folds
folds = createFolds(unique(d_train$SUBJ), 10)

# Preperations for loop - result lists and n reset
rmse_train = NULL
rmse_test = NULL
n=1

# Loop
for(i in folds){
   # Create test dataset
  test = subset(d_train,SUBJ %in% i) # dataset with one fold
  
  # Create training dataset
  train = subset(d_train,!(SUBJ %in% i)) # take all the values in SUBJ and tell when they are in the list of values (dataset with all other folds)
 
  # Make a model on the training dataset
  m_temp = lmer(CHI_MLU ~ 1 + VISIT + Diagnosis + (1+VISIT|SUBJ), data=train)
  
  # Check error between fit of training data and actual training data
  rmse_train[n] = Metrics :: rmse(train$CHI_MLU, fitted(m_temp))
  
  # Check error between predicitions for test data and actual test data
  rmse_test[n] = Metrics :: rmse(test$CHI_MLU, predict(m_temp, test, allow.new.levels=T)) 

    # Loop end and n+1
  n=n+1
}

# Get results
rmse_train
mean(rmse_train)

rmse_test
mean(rmse_test)

```
- Report the results and comment on them.
Cross-validation on 10 folds showed that the basic model (Child MLU predicted by visit and diagnosis with by-subject random slopes and intercepts for visits as random effects) had a mean rmse of 0.3410 on the training data and 0.8026 on the test data.

- Now try to find the best possible predictive model of ChildMLU, that is, the one that produces the best cross-validated results.

---MODEL 1---
```{r}
# Create folds
folds = createFolds(unique(d_train$SUBJ), 10)

# create list
newpred = list()

# Preperations for loop - result lists and n reset
rmse_train = NULL
rmse_test = NULL
n=1

# Loop
for(i in folds){
  # Create training dataset
  train = subset(d_train,!(SUBJ %in% i)) # take all the values in SUBJ and tell when they are in the list of values
  
  # Create test dataset
  test = subset(d_train,SUBJ %in% i)
  
  # Make a model on the training dataset
  m1 = lmer(CHI_MLU ~ 1 + Diagnosis + verbalIQ + (1+VISIT|SUBJ), data=train)
  
  # Check error between fit of training data and actual training data
  rmse_train[n] = Metrics :: rmse(train$CHI_MLU, fitted(m1))
  
  # Check error between predicitions for test data and actual test data
  rmse_test[n] = Metrics :: rmse(test$CHI_MLU, predict(m1, test, allow.new.levels=T)) 

    # Loop end and n+1
  n=n+1
}

# Get results
rmse_train
mean(rmse_train)

rmse_test
mean(rmse_test)

```
rmse_train = 0.3405, rmse_test = 0.7926

---MODEL 2---
```{r}
# Create folds
folds = createFolds(unique(d_train$SUBJ), 10)

# create list
newpred = list()

# Preperations for loop - result lists and n reset
rmse_train = NULL
rmse_test = NULL
n=1

# Loop
for(i in folds){
  # Create training dataset
  train = subset(d_train,!(SUBJ %in% i)) # take all the values in SUBJ and tell when they are in the list of values
  
  # Create test dataset
  test = subset(d_train,SUBJ %in% i)
  
  # Make a model on the training dataset
  m2 = lmer(CHI_MLU ~ 1 + VISIT  + Diagnosis + verbalIQ + (1+VISIT|SUBJ), data=train)
  
  # Check error between fit of training data and actual training data
  rmse_train[n] = Metrics :: rmse(train$CHI_MLU, fitted(m2))
  
  # Check error between predicitions for test data and actual test data
  rmse_test[n] = Metrics :: rmse(test$CHI_MLU, predict(m2, test, allow.new.levels=T)) 

    # Loop end and n+1
  n=n+1
}

# Get results
rmse_train
mean(rmse_train)

rmse_test
mean(rmse_test)

```
rmse_train = 0.3469, rmse_test = 0.6411

---MODEL 3---
```{r}
# Create folds
folds = createFolds(unique(d_train$SUBJ), 10)

# create list
newpred = list()

# Preperations for loop - result lists and n reset
rmse_train = NULL
rmse_test = NULL
n=1

# Loop
for(i in folds){
  # Create training dataset
  train = subset(d_train,!(SUBJ %in% i)) # take all the values in SUBJ and tell when they are in the list of values
  
  # Create test dataset
  test = subset(d_train,SUBJ %in% i)
  
  # Make a model on the training dataset
  m3 = lmer(CHI_MLU ~ 1 + VISIT + Diagnosis + verbalIQ + types_CHI + (1+VISIT|SUBJ), data=train)
  
  # Check error between fit of training data and actual training data
  rmse_train[n] = Metrics :: rmse(train$CHI_MLU, fitted(m3))
  
  # Check error between predicitions for test data and actual test data
  rmse_test[n] = Metrics :: rmse(test$CHI_MLU, predict(m3, test, allow.new.levels=T)) 

    # Loop end and n+1
  n=n+1
}

# Get results
rmse_train
mean(rmse_train)

rmse_test
mean(rmse_test)

```
rmse_train = 0.3056, rmse_test = 0.4671  - THE WINNER

---MODEL 4---
```{r}
# Create folds
folds = createFolds(unique(d_train$SUBJ), 10)

# create list
newpred = list()

# Preperations for loop - result lists and n reset
rmse_train = NULL
rmse_test = NULL
n=1

# Loop
for(i in folds){
  # Create training dataset
  train = subset(d_train,!(SUBJ %in% i)) # take all the values in SUBJ and tell when they are in the list of values
  
  # Create test dataset
  test = subset(d_train,SUBJ %in% i)
  
  # Make a model on the training dataset
  m4 = lmer(CHI_MLU ~ 1 + VISIT + Diagnosis + verbalIQ + types_CHI + tokens_CHI + (1+VISIT|SUBJ), data=train)
  
  # Check error between fit of training data and actual training data
  rmse_train[n] = Metrics :: rmse(train$CHI_MLU, fitted(m4))
  
  # Check error between predicitions for test data and actual test data
  rmse_test[n] = Metrics :: rmse(test$CHI_MLU, predict(m4, test, allow.new.levels=T)) 

    # Loop end and n+1
  n=n+1
}

# Get results
rmse_train
mean(rmse_train)

rmse_test
mean(rmse_test)

```
rmse_train = 0.3013, rmse_test = 0.4695

---MODEL 5---
```{r}
# Create folds
folds = createFolds(unique(d_train$SUBJ), 10)

# create list
newpred = list()

# Preperations for loop - result lists and n reset
rmse_train = NULL
rmse_test = NULL
n=1

# Loop
for(i in folds){
  # Create training dataset
  train = subset(d_train,!(SUBJ %in% i)) # take all the values in SUBJ and tell when they are in the list of values
  
  # Create test dataset
  test = subset(d_train,SUBJ %in% i)
  
  # Make a model on the training dataset
  m5 = lmer(CHI_MLU ~ 1 + VISIT + Diagnosis + verbalIQ + types_CHI + tokens_CHI + nonVerbalIQ + (1+VISIT|SUBJ), data=train)
  
  # Check error between fit of training data and actual training data
  rmse_train[n] = Metrics :: rmse(train$CHI_MLU, fitted(m5))
  
  # Check error between predicitions for test data and actual test data
  rmse_test[n] = Metrics :: rmse(test$CHI_MLU, predict(m5, test, allow.new.levels=T)) 

    # Loop end and n+1
  n=n+1
}

# Get results
rmse_train
mean(rmse_train)

rmse_test
mean(rmse_test)

```
rmse_train = 0.3018, rmse_test = 0.4728

The model that was best in prediction was the one we chose in the previous assignment (diagnosis, verbal IQ, and types_CHI as fixed effects, and by-subject random slopes and intercepts for visits as random effects). Mean rmse on the training data was 0.3056 and 0.4671 on the test data.

- Bonus Question 1: How would you go comparing the performance of the basic model and the cross-validated model on the testing set?
- Bonus Question 2: What is the effect of changing the number of folds? Can you plot RMSE as a function of number of folds?
- Bonus Question 3: compare the cross-validated predictive error against the actual predictive error on the test data


### Exercise 3) Assessing the single child

Let's get to business. This new kiddo - Bernie - has entered your clinic. This child has to be assessed according to his group's average and his expected development.

Bernie is one of the six kids in the test dataset, so make sure to extract that child alone for the following analysis.
```{r}
# Create the final model
m_final = lmer(CHI_MLU ~ 1 + VISIT + Diagnosis + types_CHI + verbalIQ + (1+VISIT|SUBJ), data=train)

# Create Bernie subset (kid numero 2)
bernie = subset(d_test,SUBJ==2)
```

You want to evaluate:

- how does the child fare in ChildMLU compared to the average TD child at each visit? Define the distance in terms of absolute difference between this Child and the average TD.
(Tip: recreate the equation of the model: Y=Intercept+BetaX1+BetaX2, etc; input the average of the TD group  for each parameter in the model as X1, X2, etc.).

```{r}
#Get average verbalIQ for TD kids (subset for kids with TD, choose verbalIQ, get mean)
avg_verbalIQ = mean(subset(d_train, Diagnosis == "TD")$verbalIQ)
avg_CHI_MLU = mean(subset(d_train, Diagnosis == "TD")$CHI_MLU)
avg_ADOS = mean(subset(d_train, Diagnosis == "TD")$ADOS)
avg_MOT_MLU = mean(subset(d_train, Diagnosis == "TD")$MOT_MLU)
avg_types_CHI = mean(subset(d_train, Diagnosis == "TD")$types_CHI)

# Create fake average kid dataset
Avg = data.frame(SUBJ = rep(666, 6),
                    Diagnosis = rep("TD", 6),
                    VISIT = c(1:6), 
                    verbalIQ = rep(avg_verbalIQ, 6),
                    ADOS = rep(avg_ADOS, 6),
                    MOT_MLU = rep(avg_MOT_MLU, 6),
                    types_CHI = rep(avg_types_CHI, 6)
                    )

# Predict Avg's scores
Avg_scores = predict(m_final, Avg, allow.new.levels=T)

# Check numerical difference between Bernie and Avg
   # - Plot them together
plot(bernie$CHI_MLU, ylim = c(1,4)) +
lines(Avg_scores)

# Define absolute distance between Bernie and Avg
bernie$CHI_MLU - Avg_scores

```
Bernie did better than average TD child in all visits except the first one. The absolute distances between Bernie and average TD were the following: -0.0683 (visit1), 0.4054 (visit2), 1.1279 (visit3), 0.8716 (visit4), 0.7755 (visit5), 0.9644 (visit6).


- how does the child fare compared to the model predictions at Visit 6? Is the child below or above expectations? (tip: use the predict() function on Bernie's data only and compare the prediction with the actual performance of the child)

```{r}
# Bernie's actual MLU at visit 6
bernie$CHI_MLU[6]
  # 3.448

# 6th visits is no. 11 in this case
# Bernie's predicted MLU
predict(mb,bernie)
  # 3.956 for basic model
predict(m_final,bernie)
  # 3.468 for cross-validated model
```
Bernie's real score on the 6th visit (3.448) was close to the model's predicted value (3.468), although the model slightly overestimated child’s performance.



### OPTIONAL: Exercise 4) Model Selection via Information Criteria
Another way to reduce the bad surprises when testing a model on new data is to pay close attention to the relative information criteria between the models you are comparing. Let's learn how to do that!

Re-create a selection of possible models explaining ChildMLU (the ones you tested for exercise 2, but now trained on the full dataset and not cross-validated).

Then try to find the best possible predictive model of ChildMLU, that is, the one that produces the lowest information criterion.

- Bonus question for the optional exercise: are information criteria correlated with cross-validated RMSE? That is, if you take AIC for Model 1, Model 2 and Model 3, do they co-vary with their cross-validated RMSE?

### OPTIONAL: Exercise 5): Using Lasso for model selection
Welcome to the last secret exercise. If you have already solved the previous exercises, and still there's not enough for you, you can expand your expertise by learning about penalizations. Check out this tutorial: http://machinelearningmastery.com/penalized-regression-in-r/ and make sure to google what penalization is, with a focus on L1 and L2-norms. Then try them on your data!


